---
title: "Calibrating per-sample gene set scores"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibrating per-sample gene set scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What this package is for

A common workflow in cancer genomics: compute a per-sample score for each gene
set (ssGSEA, GSVA, or a mean of standardized expression), then treat that score
as an ordinary variable — compare it between groups, put it in a Cox model, test
it against stage or survival. The p-values that come out assume the genes in the
set are independent. They are not.

This package computes an empirical p-value instead, using a null distribution
built from random gene sets matched on size and mean expression. It also tells
you when that null cannot be trusted.

## A worked example

Simulate an expression matrix where nothing is happening.

```{r}
library(gscalibrate)
set.seed(1)

n_genes <- 3000; n_samples <- 80
expr <- matrix(rnorm(n_genes * n_samples), n_genes, n_samples,
               dimnames = list(paste0("gene", seq_len(n_genes)),
                               paste0("s", seq_len(n_samples))))
group <- rep(c(TRUE, FALSE), each = n_samples / 2)

sets <- list(setA = paste0("gene", 1:120),
             setB = paste0("gene", 201:320))

calibrate_geneset(expr, sets, group, n_null = 100)
```

With independent genes and no effect, nominal and empirical p-values agree and
`floor_std` sits near the theoretical 1.96. `ratio` is `TRUE`: the tested sets
are no more internally correlated than a random draw.

## Where the parametric p-value goes wrong

Now give the whole transcriptome a mild shift between groups — as happens
whenever the grouping variable tracks something real in the tumour — and make
one set internally coherent, as real gene sets are.

```{r}
set.seed(2)
expr2 <- expr

# 40% of all genes differ slightly between groups
de <- sample(n_genes, 0.4 * n_genes)
expr2[de, group] <- expr2[de, group] + rnorm(length(de), 0, 0.3)

# setA shares a latent factor: a coherent set
lat <- rnorm(n_samples)
idx <- match(sets$setA, rownames(expr2))
expr2[idx, ] <- sqrt(0.15) * matrix(rep(lat, each = length(idx)),
                                    length(idx), n_samples) +
                sqrt(0.85) * expr2[idx, ]

res <- calibrate_geneset(expr2, sets, group, n_null = 100)

        
```

Two things to read here.

`floor_std` is the 95th percentile of the null coefficient expressed in units of
its own standard error. Compare it to **1.96**. When it is much larger, the
parametric test is miscalibrated for this data and this grouping — a nominal
p-value of 0.01 does not mean what it usually means.

`ratio` is `FALSE` for `setA`. Its internal correlation (`rho_set`) exceeds
anything a random draw achieved (`sigma_within`), so the null was built from sets
less coherent than the one being tested, and the empirical p-value is itself
anti-conservative.

## The limitation, stated plainly

A draw-based null cannot be fixed by matching draws on correlation. In TCGA lung
adenocarcinoma, Human Protein Atlas tissue programs reach mean inter-gene
correlation of 0.245, while random draws of the same size reach at most 0.021.
There is no candidate pool to draw a match from, because coherence is what makes
a gene set a gene set rather than a list.

Simulated type I error for a coherent set (ρ = 0.15 against a background of 0.05)
is 0.125 at 250 genes and 0.255 at 900 genes, against a nominal 0.05.

So when `ratio` is `FALSE`, this package is not the right tool. Use one that
estimates correlation from the set itself:

```{r, eval = FALSE}
library(limma)
design <- model.matrix(~ group)
roast(expr2, index = which(rownames(expr2) %in% sets$setA),
      design = design, contrast = 2)
cameraPR(statistic, index)   # direct replacement for pre-ranked GSEA
```

In the same simulation, `roast` gives type I error 0.035 at 250 genes and 0.065
at 900, with power 0.76 and 0.72. It is also faster. Note that `roast` itself
over-rejects for small sets (0.135 at 75 genes), so no method here is universally
correct.

## When this package is still useful

Three cases.

**Diagnosis.** `floor_std` quantifies how far the parametric null is from theory
for your particular data and grouping. That is worth knowing before you trust any
p-value from this workflow, whichever test produced it.

**Weakly coherent sets.** When `ratio` is `TRUE`, the empirical p-value is
sound and the procedure is straightforward to explain.

**Auditing published results.** Applied to a pre-registered pan-cancer analysis,
28 of 43 associations significant after Benjamini–Hochberg correction — 65% —
failed empirical calibration. Independently, `limma::camera` agreed with the
empirical null on 90 of 99 tests with no contradictions in either direction.

## Choosing `n_null`

The resolution of `p_empirical` is `1/n_null`. The default of 100 is enough to
distinguish 0.5 from 0.05 but not 0.07 from 0.13. When a result sits near your
threshold, raise it to 1000.

## Reference

Wu D, Smyth GK (2012). Camera: a competitive gene set test accounting for
inter-gene correlation. *Nucleic Acids Research* 40, e133.
