---
title: "Calibrating per-sample gene set scores"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Calibrating per-sample gene set scores}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What this package is for

A common workflow in cancer genomics: compute a per-sample score for each gene
set (ssGSEA, GSVA, or a mean of standardized expression), then treat that score
as an ordinary variable — compare it between groups, put it in a Cox model, test
it against stage or survival.

The p-value that comes out answers a **self-contained** question: is this set
associated with the outcome at all? That is usually not the question people mean
to ask. They mean the **competitive** question: is this set *more* associated
than a typical gene set would be? The distinction is due to Goeman and Bühlmann
(2007) and it matters enormously when many genes respond to the grouping.

This package computes a competitive comparison against random gene sets matched
on size and mean expression, and reports how far that null sits from the
parametric one.

## Why the distinction matters

In TCGA lung adenocarcinoma, comparing LKB1-deficient with intact tumours, **55%
of all genes differ individually at p < 0.05.** With background activity that
high, essentially every gene set contains active genes. Running `rSEA` on the 50
Hallmark sets there gives **50 of 50 self-contained significant** (adjusted p
between 10⁻¹¹ and 10⁻⁵¹) and **9 of 50 competitively significant**.

The self-contained answer is correct and useless. The competitive answer is the
interesting one, and it is much harder to get right.

## A worked example

```{r}
library(gscalibrate)
set.seed(1)

n_genes <- 3000; n_samples <- 80
expr <- matrix(rnorm(n_genes * n_samples), n_genes, n_samples,
               dimnames = list(paste0("gene", seq_len(n_genes)),
                               paste0("s", seq_len(n_samples))))
group <- rep(c(TRUE, FALSE), each = n_samples / 2)
sets  <- list(setA = paste0("gene", 1:120), setB = paste0("gene", 201:320))

calibrate_geneset(expr, sets, group, n_null = 100)
```

With independent genes and no effect, nominal and empirical p-values agree and
`floor_std` sits near the theoretical 1.96.

## Reading the output

`p_nominal` is the self-contained p-value from the model.

`p_empirical` is the competitive comparison against matched-random sets.

`floor_std` is the 95th percentile of the null coefficient in units of its own
standard error. **Compare it to 1.96.** Values well above that mean the
parametric null is badly miscalibrated for this data and grouping — in real
TCGA cohorts it commonly runs 5 to 7.

`rho_set` is the set's mean inter-gene correlation, computed on residuals after
removing the group effect so that a real difference does not inflate it.

## Survival models

The workflow this package was written about most often uses Cox regression, so
that is supported directly:

```{r, eval = FALSE}
calibrate_geneset(expr, sets, model = "cox",
                  time = clin$os_time, event = clin$os_event,
                  covariates = data.frame(purity = clin$purity))
```

## An important limitation

**`p_empirical` is a diagnostic, not a corrected p-value.**

The null is computed within one fixed grouping, so it captures variation from
which genes were drawn and not from which samples fall in each group. For a real
gene set the second component is larger. Measured on HALLMARK_E2F_TARGETS in
TCGA LUAD: the within-grouping null has SD 0.047 while the set's own coefficient
varies across random sample splits with SD 0.110. Against random splits, where
the null is true by construction, type I error reaches 0.30 for some sets.

Two repairs were tested and neither works:

- **Matching draws on inter-gene correlation is impossible.** Real tissue
  programs reach mean correlation 0.245 while random draws of the same size
  reach at most 0.021. There is no candidate pool to match from, because
  coherence is what makes a gene set a gene set rather than a list.
- **Combining gene randomization with sample permutation reverts to the
  self-contained null**, a result due to Maciejewski (2014) and confirmed here:
  a set with competitive p = 0.110 gave restandardized p = 0.003, tracking its
  nominal p of 5×10⁻¹³.

For inference under dependence, use a method built for it:

```{r, eval = FALSE}
limma::camera(expr, index, design)      # estimates inter-gene correlation
limma::roast(expr, index, design)       # rotation, preserves correlation
rSEA::SEA(pvalue, featureIDs, pathlist) # closed testing, valid under dependence
```

These three do not always agree. In TCGA LUAD across 50 Hallmark sets, 38 are
significant after Benjamini–Hochberg alone, 6 survive the empirical null, 9
survive rSEA's competitive test, and only 4 survive both. **Benjamini–Hochberg
alone keeps 38 of 50; any method accounting for dependence keeps 6 to 9. Which 6
to 9 depends on the method.**

## Where this package is still useful

**Diagnosis.** `floor_std` quantifies how far the parametric null is from theory
for your particular data and grouping. Worth knowing before trusting any p-value
from this workflow, whichever test produced it.

**Auditing.** Applied to a pre-registered pan-cancer analysis, 28 of 43
associations significant after Benjamini–Hochberg failed empirical calibration.
Across four cohorts and 50 Hallmark sets, 84 of 105 failed. `limma::camera`
independently agreed with the empirical null on 90 of 99 tests with no
contradictions in either direction.

## Choosing `n_null`

Resolution of `p_empirical` is `1/(n_null + 1)`. The default of 100 distinguishes
0.5 from 0.05 but not 0.07 from 0.13. Raise it to 1000 when a result sits near
your threshold.

## References

Goeman JJ, Bühlmann P (2007). Analyzing gene expression data in terms of gene
sets: methodological issues. *Bioinformatics* 23, 980–987.

Wu D, Smyth GK (2012). Camera: a competitive gene set test accounting for
inter-gene correlation. *Nucleic Acids Research* 40, e133.

Maciejewski H (2014). Gene set analysis methods: statistical models and
methodological differences. *Briefings in Bioinformatics* 15, 504–518.

Ebrahimpoor M, Spitali P, Hettne K, Tsonaka R, Goeman J (2020). Simultaneous
enrichment analysis of all possible gene-sets. *Briefings in Bioinformatics* 21,
1302–1312.

Phipson B, Smyth GK (2010). Permutation P-values should never be zero.
*Statistical Applications in Genetics and Molecular Biology* 9, 39.
